Abstraction Refinement for Emptiness Checking of Alternating Data Automata

### 2017-10-21 tacas/draft.tex
experiment result
ok

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) using only atoms $x \geq k$
and $x = \overline{x} + k$, for $k \in \zed$, with the classical
interpretation of the function symbols on integers. Since reachability
of a control state is undecidable for AVASS \cite{LINCOLN92}, 
emptiness is undecidable for ADA.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{cc}
\includegraphics[scale=0.5]{PA1.pdf} & \includegraphics[scale=0.5]{PA2.pdf} \\[-2mm]
\tiny{(a)} & \tiny{(b)} \\
\\
\includegraphics[scale=0.5]{PA3.pdf} & \includegraphics[scale=0.5]{PA4.pdf} \\[-2mm]
\tiny{(c)} & \tiny{(d)}
\end{tabular}
\end{center}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

  \If{$\Lambda(n_i) \not\models I_i$}

@@

\If{$\Lambda(x) \models \Lambda(y)$}

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\begin{tabular}{c}
\includegraphics[scale=0.60]{IM1.pdf} \\
\tiny{(a)} \\
\includegraphics[scale=0.60]{IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.60]{IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\end{minipage}
\begin{minipage}{6cm}
\includegraphics[scale=0.60]{IM4.pdf}
\centerline{\tiny{(d)}}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact}}
\label{fig:im}
\end{figure}

@@

@@

We have implemented both Algorithm \ref{alg:predabs} and
\ref{alg:impact} in a prototype tool\footnote{The implementation is
  available at \url{https://github.com/cathiec/AltImpact}} that uses
the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} for
the satisfiability queries and interpolant generation, in the theory
of linear integer arithmetic (LIA) combined with booleans. We compared
both algorithms with a previous implementation of a trace inclusion
procedure, called
\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
that uses on-the-fly determinisation and lazy predicate abstraction
with interpolant-based refinement \cite{Tacas16} in the LIA theory,
without booleans.

The results of the experiments are given in Table
\ref{tab:experiments}. We applied the tool first to several array
logic entailments, which occur as verification conditions for
imperative programs with arrays \cite{cav09} (array\_shift,
array\_simple, array\_rotation 1+2) available online
\cite{ntslib}. Next, we applied it on proving safety properties of
hardware circuits (hw1+2) \cite{smrcka}. Finally, we considered two
timed communication protocols, consisting of systems that are
asynchronous compositions of timed automata, whom correctness
specifications are given by timed automata monitors: a timed version of
the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
results were obtained on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz
with 8GB of RAM. The automata sizes are in bytes and the execution
times are in seconds.

\begin{table}[htb]
\begin{center}
{\fontsize{8}{9}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
\hline
simple1 & 309 & no & 0.641 & 0.061 & 0.021\\
\hline
simple2 & 504 & yes & 0.667 & 0.179 & 0.031\\
\hline
simple3 & 214 & yes & 0.825 & 0.150 & 0.051\\
\hline
array\_shift & 874 & yes & 2.138 & 0.209 & 0.060\\
\hline
array\_simple & 3440 & yes & timeout & 16.435 & 5.496\\
\hline
array\_rotation1 & 1834 & yes & 6.827 & 0.865 & 0.114\\
\hline
array\_rotation2 & 15182 & yes & timeout & timeout & 23.056\\
\hline
abp & 6909 & no & 8.292 & 1.410 & 1.715\\
\hline
train & 1823 & yes & 16.989 & 2.738 & 0.308\\
\hline
hw1 & 322 & yes & 1.566 & 0.298 & 0.159\\
\hline
hw2 & 674 & yes & 22.914 & 0.552 & 0.419\\
\hline
\end{tabular}
}
\caption{}\label{tab:experiments}
\end{center}
\end{table}

As in the non-deterministic case \cite{mcmillan06}, \impact~
outperforms lazy predicate abstraction for checking emptiness by at
least one order of magnitude. However, both our implementations are
slower than \textsc{Includer}, on average (except for the abp
example). The reason for this is currently under investigation, one
possible bottleneck being the hardness of the combined (LIA+booleans)
interpolation problems, as opposed to converting the entire formula
into DNF, eliminating the boolean variables and using interpolation in
the pure LIA theory.

>>>>

The emptiness problem for ADA is undecidable, even in very simple
cases. For instance, if $\Data^\I$ is the set of positive integers, an
ADA can simulate an Alternating Vector Addition System with States
(AVASS) using only atoms $x \geq k$ and $x = \overline{x} + k$, for $k
\in \zed$, with the classical interpretation of the function symbols
on integers. Since reachability of a control state is undecidable for
AVASS \cite{LINCOLN92}, ADA emptiness is undecidable.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{cc}
\includegraphics[scale=0.5]{PA1.pdf} & \includegraphics[scale=0.5]{PA2.pdf} \\[-2mm]
\tiny{(a)} & \tiny{(b)} \\
\\
\includegraphics[scale=0.5]{PA3.pdf} & \includegraphics[scale=0.5]{PA4.pdf} \\[-2mm]
\tiny{(c)} & \tiny{(d)}
\end{tabular}
\end{center}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
\label{fig:pa}
\end{figure}

@@

  \If{$\Lambda(n_i) \not\models I_i$}
  \label{ln:entailment1}

@@

\If{$\Lambda(x) \models \Lambda(y)$}
\label{ln:entailment2}

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\begin{tabular}{c}
\includegraphics[scale=0.60]{IM1.pdf} \\
\tiny{(a)} \\
\includegraphics[scale=0.60]{IM2.pdf} \\[-2mm]
\tiny{(b)} \\
\includegraphics[scale=0.60]{IM3.pdf} \\[-2mm]
\tiny{(c)}
\end{tabular}
\end{minipage}
\begin{minipage}{6cm}
\includegraphics[scale=0.60]{IM4.pdf}
\centerline{\tiny{(d)}}
\end{minipage}
\end{center}
\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
  by Algorithm \ref{alg:impact}}
\label{fig:im}
\end{figure}

@@

The correctness of Algorithm \ref{alg:impact} is coined by the theorem
below:

@@

We have implemented both Algorithm \ref{alg:predabs} and
\ref{alg:impact} in a prototype tool\footnote{The implementation is
  available at \url{https://github.com/cathiec/JAltImpact}} that uses
the MathSAT5 SMT solver\footnote{\url{http://mathsat.fbk.eu/}} via the
Java SMT
interface\footnote{\url{https://github.com/sosy-lab/java-smt}} for the
satisfiability queries and interpolant generation, in the theory of
linear integer arithmetic with uninterpreted boolean functions
(UFLIA). We compared both algorithms with a previous implementation of
a trace inclusion procedure, called
\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
that uses on-the-fly determinisation and lazy predicate abstraction
with interpolant-based refinement \cite{Tacas16} in the LIA theory.

The results of the experiments are given in Table
\ref{tab:experiments}. We applied the tool first to several array
logic entailments, which occur as verification conditions for
imperative programs with arrays \cite{cav09} (array\_shift,
array\_simple, array\_rotation1+2) available online
\cite{ntslib}. Next, we applied it on proving safety properties of
hardware circuits (hw1+2) \cite{smrcka}. Finally, we considered two
timed communication protocols, consisting of systems that are
asynchronous compositions of timed automata, whom correctness
specifications are given by timed automata monitors: a timed version
of the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
results were obtained on x86\_64 Linux Ubuntu virtual machine with 8GB
of RAM running on an Intel(R) Xeon(R) CPU E5-2683 v3 @ 2.00GHz. The
automata sizes are given in bytes needed to store their ASCII
description on file and the execution times are in seconds.

\begin{table}[htb]
\vspace*{-\baselineskip}
\begin{center}
{\fontsize{8}{9}\selectfont
\begin{tabular}{||l|c|c|c|c|c||}
\hline
Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} (sec) & Algorithm \ref{alg:impact} (sec) & \textsc{Includer} (sec) \\
\hline
simple1 & 309 & no & 0.774 & 0.064 & 0.076\\
\hline
simple2 & 504 & yes & 0.867 & 0.070 & 0.070\\
\hline
simple3 & 214 & yes & 0.899 & 0.095 & 0.095\\
\hline
array\_shift & 874 & yes & 2.889 & 0.126 & 0.078\\
\hline
array\_simple & 3440 & yes & timeout & 9.998 & 7.154\\
\hline
array\_rotation1 & 1834 & yes & 7.227 & 0.331 & 0.229\\
\hline
array\_rotation2 & 15182 & yes & timeout & timeout & 31.632\\
\hline
abp & 6909 & no & 9.492 & 0.631 & 2.288\\
\hline
train & 1823 & yes & 19.237 & 0.763 & 0.678\\
\hline
hw1 & 322 & yes & 1.861 & 0.163 & 0.172\\
\hline
hw2 & 674 & yes & 24.111 & 0.308 & 0.473\\
\hline
\end{tabular}
}
\caption{}\label{tab:experiments}
\end{center}
\vspace*{-2\baselineskip}
\end{table}

As in the case of non-alternating nondeterministic integer programs
\cite{mcmillan06}, the alternating version of \impact~ (Algorithm
\ref{alg:impact}) outperforms lazy predicate abstraction for checking
emptiness by at least one order of magnitude. Moreover, \impact~ is
comparable, on average, to the previous implementation of
\textsc{Includer}, which uses also MathSAT5 via the C API. We believe
the reason for which \textsc{Includer} outperforms \impact~ on some
examples is the hardness of the UFLIA entailment checks used in
Algorithm \ref{alg:impact} (line \ref{ln:entailment1} and line
\ref{ln:entailment2} in the function \Call{Close}{}) as opposed to the
pure LIA entailment checks used in \textsc{Includer}. According to our
statistics, Algorithm \ref{alg:impact} spends more than 50\% of the
time waiting for the SMT solver to finish answering entailment
queries.
